---
title: "'HiGHS' optimization solver with OMPR algebraic modeling system: A Small Example"
format: html
date: September 4, 2022
author: notesofdabbler
---

```{r}
# load libraries and source functions
library(highs)
library(ompr)
library(ompr.roi)
library(dplyr)

source("ompr_helperfns.R")
```

We show an example of using [OMPR](https://dirkschumacher.github.io/ompr/) alebraic modeling system for optimization problems with [HiGHS](https://highs.dev/) solver using R package [highs](https://cran.r-project.org/web/packages/highs/index.html). Only portions of R code are shown here. The `.qmd` with all embedded R code is in this [location](https://github.com/notesofdabbler/optwithR/blob/main/highs_example_ompr.qmd).

:::: {.columns}

::: {.column width="40%"}

$$
\begin{array}{ll}
\min & x_0 + x_1 + 3 \\
& x_1 \leq 7 \\
& 5 \leq x_0 + 2x_1 \leq 15 \\
& 6 \leq 3x_0 + 2x_1 \\
& 0 \leq x_0 \leq 4 \\
& 1 \leq x_1
\end{array}
$$
:::

::: {.column width="60%"}

```{r}
mdl = MIPModel() %>%
      add_variable(x0, lb = 0, ub = 4, type = "continuous") %>%
      add_variable(x1, lb = 1, type = "continuous") %>%
      set_objective(x0+x1+3) %>%
      add_constraint(x1 <= 7) %>%
      add_constraint(x0 + 2*x1 <= 15) %>%
      add_constraint(x0 + 2*x1 >= 5) %>%
      add_constraint(3*x0 + 2*x1 >= 6)
```

:::
::::

OMPR is integrated with [ROI](https://cran.r-project.org/web/packages/ROI/index.html) package to solve the model. However, currently I didn't see a ROI plugin for highs solver. So I adapted the `as_ROI_model` function in package `ompr.ROI` to get the needed elements to pass to highs solver.

```{r}
highs_mdl = as_highs_model(mdl)

s <- highs_solve(L = as.numeric(highs_mdl$L), lower = highs_mdl$lower, upper = highs_mdl$upper,
                 A = highs_mdl$A, lhs = highs_mdl$lhs, rhs = highs_mdl$rhs,
                 offset = highs_mdl$offset)

s[["status"]]
s[["objective_value"]]
s[["primal_solution"]]
```